---
title: "Assignment 2 - Data Cleaning"
author: "Sam Muir"
format: html
---

```{r}
#| message: false  

# Libraries
library(tidyverse)

# file names
datadir_raw <- "data/raw/"

datadir_processed <- "data/processed/"

species_file <- "ASDN_Daily_species.csv"

snowsurvey_file <- "ASDN_Snow_survey.csv"
```

```{r}
# Import the species
snowsurvey_csv <- read_csv("data/snow_cover_fixed.csv")

glimpse(snowsurvey_csv)
```

### Cleaning in `Snow_cover` column
```{r}
snowsurvey_fixed <- snowsurvey_csv %>%
  mutate(Snow_cover = ifelse(Snow_cover %in% c(".", "-", "n/a", "unk"), NA, Snow_cover))

snowsurvey_fixed <- snowsurvey_fixed %>%
  mutate(Snow_cover = ifelse(Snow_cover == "<1", 0, Snow_cover))

snowsurvey_fixed <- snowsurvey_fixed %>%
  mutate(Snow_cover = as.numeric(Snow_cover))

snowsurvey_fixed <- snowsurvey_fixed %>%
  # filter(Snow_cover > 100) %>%
  mutate(Snow_cover = ifelse(Snow_cover > 100, NA, Snow_cover))

snowsurvey_fixed %>%
  filter(Snow_cover < 0)
```


### Cleaning the `Water_cover` column
```{r}
snowsurvey_fixed %>%
  count(Water_cover) %>%
  filter(is.na(as.numeric(Water_cover)))
```
- We have values that are not numeric: "n/a", "unk", "-", and "."

```{r}
# turning these character values in NAs
# from the Readme: the "-" indicates an absence of data
watercover_fixed <- snowsurvey_fixed %>%
  mutate(Water_cover = ifelse(Water_cover %in% c("n/a", "unk", "-", "."), NA, Water_cover),
         Water_cover = as.numeric(Water_cover))

glimpse(watercover_fixed)
```

```{r}
# check for values over 100
watercover_fixed %>%
  filter(Water_cover > 100)

watercover_fixed <- watercover_fixed %>%
  mutate(Water_cover = ifelse(Water_cover > 100, NA, Water_cover))
```

- In this case we are assuming that the water cover of 353 is incorrect. This row also has an illogical value of -298 for land cover. Though the value for snow cover does make sense (45). We could assume that the 45 is correct and that with another combination of Snow and Water cover will equal a total cover of 100. But, I think it makes more sense to assume that all of the values in that row are incorrect and make everything NA. Even though 45 is a reasonable value, it still could be a mistakenly recorded values like the others in the row.  

```{r}
# check for negative values
watercover_fixed %>%
  filter(Water_cover < 0)
# no negative values
```

### Cleaning the `Land_cover` column
```{r}
watercover_fixed %>%
  count(Land_cover) %>%
  filter(is.na(as.numeric(Land_cover)))
```
- We have values that are not numeric: "n/a", "unk", "-", and "."

```{r}
# turning these character values in NAs
# from the Readme: the "-" indicates an absence of data
landcover_fixed <- watercover_fixed %>%
  mutate(Land_cover = ifelse(Land_cover %in% c("n/a", "unk", "-", "."), NA, Land_cover),
         Land_cover = as.numeric(Land_cover))

glimpse(landcover_fixed)
```

```{r}
# check for values over 100
landcover_fixed %>%
  filter(Land_cover > 100)
# no values over 100

# check for negative values
landcover_fixed %>%
  filter(Land_cover < 0)
```
There are two Land_cover values that are less than 0: -100 & -298. In this case, each of these is part of a row that also have values that do not make sense for the other cover categories. I do not want to try to guess what they may have been intending, so I am going to set all of the cover type values for these two rows to NA. 

```{r}
landcover_fixed <- landcover_fixed %>%
  mutate(Snow_cover = ifelse(Land_cover < 0, NA, Snow_cover),
         Water_cover = ifelse(Land_cover < 0, NA, Water_cover),
         Land_cover = ifelse(Land_cover < 0, NA, Land_cover),
         Total_cover = ifelse(Land_cover < 0, NA, Total_cover))

landcover_fixed %>%
  filter(Land_cover < 0)
```

Looking at the new landcover_fixed data frame, I can see that there are cases where there are non-zero values entered for certain cover types, but the total cover is 0. 

### Cleaning the `Total_cover` column

```{r}
landcover_fixed %>%
  count(Total_cover) %>%
  filter(is.na(as.numeric(Total_cover)))

totalcover_fixed <- landcover_fixed %>%
  mutate(Total_cover = as.numeric(Total_cover))
```

```{r}
# check for values over 100
totalcover_fixed %>%
  filter(Total_cover > 100)
# no values over 100

# check for negative values
totalcover_fixed %>%
  filter(Total_cover < 0)
```
 - there are no total cover values less than 0, but there are many greater than 100
 
```{r}
totalcover_fixed <- totalcover_fixed %>%
  mutate(Total_cover = ifelse(Total_cover > 100, NA, Total_cover))

totalcover_fixed %>%
  filter(Total_cover > 100)

glimpse(totalcover_fixed)
```
 
##### Making sure the cover columns equal the total cover

```{r}
totalcover_fixed %>%
  mutate(match = case_when((Snow_cover + Water_cover + Land_cover) != Total_cover ~ "no")) %>%
  select(-Notes) %>%
  filter(match == "no")

totalcover_fixed %>%
  filter(Total_cover != 100)
```
- Here I can see that there are 184 rows where snow + water + land does not equal the total in the total cover column
- There are also 4,626 rows where land cover is not equal to 100

Plan:
- if `Snow_cover` + `Water_cover` + `Land_cover` = 100 then return that sum in the `Total_cover`
- if `Snow_cover` + `Water_cover` + `Land_cover` < 100 then inspect and see what's going on
- if `Snow_cover` + `Water_cover` + `Land_cover` > 100 then return NAs for all the cover types

```{r}
cover_fixed <- totalcover_fixed %>%
  mutate(total_fixed = ifelse((Snow_cover + Water_cover + Land_cover) == 100, (Snow_cover + Water_cover + Land_cover), NA))

cover_fixed %>%
  filter(Snow_cover + Water_cover + Land_cover < 100) %>%
  select(-Notes)
```

